Inference using context-dependent models of mutation via bounds on information propagation

### 2021-10-31 writeups/context-dependence.tex
working through the approx proof

For inference, we need to compute the $|\calS|^{n} \times |\calS|^m$ matrix $F$ whose $(a,b)^\text{th}$ entry is $p_{n,\ell,r}(t,a,b)$,
for various $t$.
This matrix is a projection of the $|\calS|^{n} \times |\calS|^{n}$ matrix whose $(a,c)^\text{th}$ entry is
\begin{align}
    p_{n}(a,c) = \left( e^{t G(n)} \right)_{a,c} ,
\end{align}
where $G(n)$ is the sparse $|\calS|^{n} \times |\calS|^{n}$ matrix defined in \eqref{eqn:G_defn}.
The ``projection'' we need just marginalizes over long (i.e.\ length $n$) patterns $c$ that match the shorter (i.e.\ length $m$) pattern $b$
as in \eqref{eq:pmMarg}:
if we define the $n \times m$ matrix $U$ so that $U_{cb}=1$ if $c_\ell^m=b$ and $U_{cb}=0$ otherwise,
then
\begin{align} \label{eqn:Tmer_trans}
    p_{m,\ell}(t,a,b) = \sum_{c \in \calS^{n}} \left( e^{t G(n)} \right)_{ac} U_{cb} ,
\end{align}
so in fact we don't need the entire matrix $e^{t G(n)}$,
just the product of this matrix with each of the columns of $U$.
Modern techniques in sparse matrix computation (e.g., Krylov methods) provide efficient ways to do this.

@@

First we quantify and prove the approximation \eqref{eqn:window_approx}.
The basic idea is that the probability is only in error if there is a chain of mutations
that create a dependency between the base of the T-mer and the flanking sequence.
To simplify the argument, first assume that all transition triples
change only a single, central site in the pattern,
taking into account context on each side of maximum length $w$.
(For instance, a trinucleotide model in which the flanking sites on each side
affect the mutation rate of the central site has $w=1$.)
%EM I suggest that we could re-introduce the mu^* notation again and remind people that these potential changes are uniform.
Recall that in our construction of the process we first chose the times and sites
at which potential changes might occur.
Suppose a potential change occurs at site $i-w$, at time $t$,
which therefore might result in a change in state at site $i$.
To resolve this, we need to know the outcome of any other potential changes
that might change the sequence between $i-w$ and $i + w$.
Consider those that might happen only to the right of $i$,
thus extending the sequence on which the change depends to the right.
The number of such potential changes is Poisson with mean $\mu^* w t$,
%EM Suggest "and their location is uniform..."
and will extend the amount of sequence on which the change at site $i$ depends
by a uniform random number of sites between 1 and $w$.
If one occurs, say at time $0 < s < t$,
then any further potential changes that occur between time 0 and time $s$
in the $w$ sites to the right of this older potential change
will again extend the range of dependency.
Putting this together, the amount of sequence to the right of a site
on which a mutation at that site depends
is bounded by the compound Poisson process $\sum_{k=1}^N W_k$,
where $N$ is Poisson with mean $\mu^* w t$
and each $W_k$ is independent and uniform on $\{1, 2, \ldots, w\}$.

@@

\begin{lemma}
    Let $X = \sum_{j=1}^N W_j$,
    where $N$ is Poisson with mean $\lambda$
    and each $W_j$ is independent and bounded between 0 and $w$.
    Then for $k \ge 1$,
    \begin{align}
        \P\{ X > k w \}
        \le
        \frac{\lambda^{k+1}}{(k+1)!} .
    \end{align}
\end{lemma}

@@

\paragraph{Full likelihood}
Now we use these ideas to quantify and prove approximation~\eqref{eqn:full_approx}.
For two positions $(i,s)$ and $(j,t)$ in sequence $\times$ time,
let $r(i,s;j,t) = |j-i|/|t-s|$ be the slope of the line between them,
and if $I$ and $J$ are subsets of sites, let $r(I,s;J,t) = \min\{r(i,s;j,t) : i\in I, j \in J\}$
be the minimum.
Above we have shown that there is a speed of dependency propagation $r_*$
such that if $r(I,s;J,t) > r_*$,
then $X_I(s)$ and $X_J(t)$ are nearly independent.
Now, fix $t$ and for each site $i$ let $N_i$ denote its $r_*$-neighborhood,
i.e.,
\begin{align}
%EM Before we had i,s,j,t, and now we have i,t,j,s. It doesn't make a difference, but should we choose one?
%EM Do we want to specify a max allowable range for j as well on the RHS?
    N_i = \{ (j,s) : r(i,t; j,s) \le r_*, \; 0 \le s \le t \} .
\end{align}
These are depicted in Figure~\ref{fig:neighborhoods}A.
We will write $i \sim j$ if $|i-j| \le 2 r_* t$,
or equivalently, if $N_i \cap N_j \neq \emptyset$,
and $i \nsim j$ otherwise.
We can think of $N_i$ as the parts of ancestral sequences in which a mutation might have affected the state at $(i,t)$,.
so if $i \nsim j$, then $i$ and $j$ are (approximately) independent at time $t$, given the sequence at time 0.
Concretely,
we have a small $\epsilon$ such that for any $i \nsim j$,
\begin{align*}
    &\big|
    \log( \P\{ X_i(t) = x_i \text{ and } X_j(t) = x_j \} ) \\
    &\qquad {} -
    \log( \P\{ X_i(t) = x_i \} \P\{ X_j(t) = x_j \} )
    ) \big|
    \le \epsilon .
\end{align*}

@@

>>>>

For inference, we need to compute the $|\calS|^{n} \times |\calS|^m$ matrix $F$ whose $(a,b)^\text{th}$ entry is $p_{n,\ell,r}(t,a,b)$,
for various $t$.
This matrix is a projection of the $|\calS|^{n} \times |\calS|^{n}$ matrix whose $(a,c)^\text{th}$ entry is
\begin{align}
    p_{n}(a,c) = \left( e^{t G(n)} \right)_{a,c} ,
\end{align}
where $G(n)$ is the sparse $|\calS|^{n} \times |\calS|^{n}$ matrix defined in \eqref{eqn:G_defn}.
The ``projection'' we need just marginalizes over long (i.e.\ length $n$) patterns $c$ that match the shorter (i.e.\ length $m$) pattern $b$
as in \eqref{eq:pmMarg}:
if we define the $n \times m$ matrix $U$ so that $U_{cb}=1$ if $c_\ell^m=b$ and $U_{cb}=0$ otherwise,
then
\begin{align} \label{eqn:Tmer_trans}
    p_{n,\ell,r}(t,a,b) = \sum_{c \in \calS^{n}} \left( e^{t G(n)} \right)_{ac} U_{cb} ,
\end{align}
so in fact we don't need the entire matrix $e^{t G(n)}$,
just the product of this matrix with each of the columns of $U$.
Modern techniques in sparse matrix computation (e.g., Krylov methods) provide efficient ways to do this.

@@

First we quantify and prove the approximation \eqref{eqn:window_approx}.
The basic idea is that the probability is only in error if there is a chain of mutations
that create a dependency between the base of the T-mer and the flanking sequence.
To simplify the argument, first assume that all transition triples
change only a single, central site in the pattern,
taking into account context on each side of maximum length $w$.
(For instance, a trinucleotide model in which the flanking sites on each side
affect the mutation rate of the central site has $w=1$.)
%EM I suggest that we could re-introduce the mu^* notation again and remind people that these potential changes are uniform.

As in the ``Simulation'' section, we start by considering a broader collection
of ``possible changes'' sampled from a homogeneous
Poisson process at a rate $\mu^*$ per site.
Suppose a possible change occurs at site $i-w$, at time $t$,
%EM should this be "change in rate"?
which therefore might result in a change in state at site $i$.
To resolve this, we need to know the outcome of any other potential changes
that might change the sequence between $i-w$ and $i + w$.
Consider those that might happen only to the right of $i$,
thus extending the sequence on which the change depends to the right.
The number of such potential changes have a uniform location
and their count is Poisson with mean $\mu^* w t$;
these extend the amount of sequence on which the change at site $i$ depends
%EM my understanding is that it's just uniform for a single mutation, but that's not obvious from the description.
by a uniform random number of sites between 1 and $w$.
If one occurs, say at time $0 < s < t$,
then any further potential changes that occur between time 0 and time $s$
in the $w$ sites to the right of this older potential change
will again extend the range of dependency.
Putting this together, the amount of sequence to the right of a site
on which a mutation at that site depends
is bounded by the compound Poisson process $\sum_{k=1}^N W_k$,
where $N$ is Poisson with mean $\mu^* w t$
and each $W_k$ is independent and uniform on $\{1, 2, \ldots, w\}$.

@@

\begin{lemma}
%EM Let's rename this X to ... V? We already have an X.
    Let $X = \sum_{j=1}^N W_j$,
    where $N$ is Poisson with mean $\lambda$
    and each $W_j$ is independent and bounded between 0 and $w$.
    Then for $k \ge 1$,
    \begin{align}
        \P\{ X > k w \}
        \le
        \frac{\lambda^{k+1}}{(k+1)!} .
    \end{align}
\end{lemma}

@@

\paragraph{Full likelihood}
Now we use these ideas to quantify and prove approximation~\eqref{eqn:full_approx}.
For two positions $(i,s)$ and $(j,t)$ in sequence $\times$ time,
let $r(i,s;j,t) = |j-i|/|t-s|$ be the absolute value of the
slope of the line between them,
and if $I$ and $J$ are subsets of sites, let $r(I,s;J,t) = \min\{r(i,s;j,t) : i\in I, j \in J\}$
be the minimum.
Above we have shown that there is a speed of dependency propagation $r_*$
such that if $r(I,s;J,t) > r_*$,
then $X_I(s)$ and $X_J(t)$ are nearly independent.
Now, fix $t$ and for each site $i$ let $N_i$ denote its $r_*$-neighborhood,
i.e.,
\begin{align}
    N_i = \{ (j,s) : r(i,s; j,t) \le r_*, \; j \leq N, \; 0 \le s \le t \} .
\end{align}
These are depicted in Figure~\ref{fig:neighborhoods}A.
We will write $i \sim j$ if $|i-j| \le 2 r_* t$,
or equivalently, if $N_i \cap N_j \neq \emptyset$,
and $i \nsim j$ otherwise.
We can think of $N_i$ as the parts of ancestral sequences in which a mutation might have affected the state at $(i,t)$.
So, if $i \nsim j$, then $i$ and $j$ are (approximately) independent at time $t$, given the sequence at time 0.
Concretely,
we have a small $\epsilon$ such that for any $i \nsim j$,
\begin{align*}
    &\big|
    \log( \P\{ X_i(t) = x_i \text{ and } X_j(t) = x_j \} ) \\
    &\qquad {} -
    \log( \P\{ X_i(t) = x_i \} \P\{ X_j(t) = x_j \} )
    ) \big|
    \le \epsilon .
\end{align*}

@@

%EM Would it be more direct to make a connection between
%EM \P\{ X_i(t) = y_i \given X_{A(L(i))}(0) = x_{A(L(i))} \text{ and } X_{L(i)}(t) = y_{L(i)} \} ,
%EM and
%EM p_{2\ell+1,\ell,\ell}(x_{i-3\ell}^{(4\ell+1)}, y_{i-2\ell}^{(2\ell+1)})
